\documentclass[12pt]{article}

% fonts

\usepackage[T1]{fontenc}
\usepackage[full]{textcomp}
\usepackage{newtxtext}
\usepackage{cabin} % sans serif
\usepackage[varqu,varl]{inconsolata} % sans serif typewriter
\usepackage[final,expansion=alltext]{microtype}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage[bigdelims,vvarbb]{newtxmath} % bb from STIX
\usepackage[cal=boondoxo]{mathalfa} % mathcal
\usepackage{csquotes}

\usepackage[backend=bibtex,style=numeric]{biblatex}
\bibliography{traffic} % Specify your BibTeX file

% geometry of the page

\usepackage[top=1in,
            bottom=1in,
            left=1in,
            right=1in]{geometry}


% paragraph spacing

\setlength{\parindent}{0pt}
\setlength{\parskip}{2ex plus 0.4ex minus 0.2ex}


\author{Matthew McAnear}
\title{Traffic Fatality Risk by Advanced Safety Features}

% useful packages

\begin{document}

\maketitle

\begin{abstract}
    Despite the proliferation of AI technology in passenger cars of collision detection/avoidance, lane drift systems, and
    various other safety systems, pedestrian and cyclist deaths are NOT significantly reduced by these
    interventions. 
\end{abstract}


\section{Introduction}

Traffic saftey is a major concern for all roadway users. The National Highway Traffic Safety Administration (NHTSA) reports
that, as of preliminary reports for through June of 2023, traffic accidents involving fatalities, in aggregate,
are down about 3\%. This decrease hides the true scale of the problem for non-road users, however. While the
overall change for pedestrians is also 3\%, the baseline level of risk for pedestrians is much higher than for 
passenger vehicles or even for pedalcyclists. Accidents involving pedestrians range anywhere from a 14\% to 
23\% baseline fatality rate, depending on the month in question \cite{national_highway_traffic_safety_administration_early_2024}.

Among drivers in the vehicle, these fatality rates have been decreasing over time, with cars becoming
safer for occupants of the vehicle. However, the overall fatality is influenced heavily by the mass of the 
car\cite{evans_car_1992}. This means that there are offsetting effects of both increasing vehicle mass and increasing 
vehicle saftey. For people outside of the vehicle, however, the statistics are alarming. According to Tyndall, 
"between 2010 and 2021 the number of pedestrians killed annually in collisions increased by 72\%, from 4300 to 
7400 \cite{tyndall_effect_2024}." Due to relaxed emissions standards for small trucks compared to passenger cars,
American buyer preferences shifted toward sport utility vehicles (SUVs) and trucks \cite{kovach_rise_2021}. Taken into 
account this connection between both the proliferation of large vehicles and the clear connection in fatality risk to
more massive and taller vehicles\cite{tyndall_effect_2024}, finding the factors that minimize pedestrian fatalities
is of critical importance for public health.

It is common knowledge that AI systems have become an increasing part of every day life, and passenger
vehicles are no exception. The question is if these improved safety systems are reducing the risk of fatalities
among pedestrians and cyclists. Next, assuming these new systems yield tangible improvements to pedestrian safety,
are the improvements large enough in magnitude to offset the increased risk from the size and height of vehicles?

In this paper, we focus primarily on those factors that influence the risk of fatalities in crashes involving pedestrians
and cyclists. After controlling for the effect of weather conditions,
geographic region, and crash variables such as time of day and intersection type, we estimate the effect
of advanged safety features on the risk of pedestrian and cyclist fatalities within vehicle types.


\section{Data}

Our data comes from the US National Highway Traffic Safety Administration (NHTSA). The NHTSA collects and distributes 
crash statistics on fatal accidents through FARS, the Fatal Accident Reporting System, and CRSS, the Crash Report 
Sampling System. Through these two datasets, we can investigate risk factors for fatal accidents and injuries in 
pedestrian and cyclist accidents. All files are available at 
[the NHTSA website](https://www.nhtsa.gov/file-downloads?p=nhtsa/downloads/). Several simplifying assumptions must be 
made in order to combine the two datasets - for example, the CRSS data is taken from 60 sampling locations, not 
nationwide, and weights are applied in order to estimate national level 
statistics\cite{national_highway_traffic_safety_administration_crash_nodate}.  To simplify the analysis, we exclude 
from consideration crashes involving multiple vehicles and focus solely on single-vehicle collisions with pedestrians 
and cyclists. We also further ignore the crash group, i.e. whether the driver or cyclist was at fault, what 
specific maneuver was being performed, or where the pedestrian happened to be in relation to the vehicle. These are 
important characteristics to consider, but the overall sparsity of the data makes inference in small subgroups
challenging.

% The result of this subsampling is that certain information is not available at the same level of default as the 
% fatality statistics. Namely, states in the fatal crash statistics have to be manually categorized into the 
% same regions reported the CRSS. Beyond this, the names of the columns are slightly different, but the overall 
% structure of the data is otherwise analogous. The dataset is fairly complex, spanning multiple files that cover the 
% accidents, vehicles involved, specific information about the vehicles, and a host of other components. To simplify
% the analysis, we exclude from consideration crashes involving multiple vehicles and focus solely on single-vehicle 
% collisions with pedestrians and cyclists. 

Another potential issue with the dataset is the overall level of sparsity in our features. While there is no 
missingness for important crash level statistics such as the bodyclass of the vehicle, things like the weight class
are missing and cannot be reliably imputed. Safety features are also missing for a large number of vehicles; to solve
this, we assume that any safety feature not listed as "standard" is not present. Though we could simulate this data 
conditional on other factors, this is an extension to a future analysis and out of scope for this short paper.

We make some other simplifying and data cleaning steps, namely collapsing the bodyclass vehicles into broader categories,
listed in the appendix.

\section{Methodology}

Our model will be a relatively traditional logistic regression, as we are primarily interested in inference on the 
effects of our variables of interest and so need a model that retains interpretability. 

We define our model as 

\begin{align*}
\alpha &\sim \text{Normal}(0, 3) \\
\sigma^2 &\sim \text{Inv-Gamma}(1, 1) \\
\alpha_{r} &\sim \text{Normal}(\alpha, \sigma^2) \\
\tau^2 &\sim \text{Inv-Gamma}(1, 1) \\
\beta &\sim \text{Laplace}(0, \tau^2) \\
\beta_{j} & \sim \text{Laplace}(\beta, \tau^2) \\
\theta_i &= \frac{1}{1 + e^{\alpha_{r_i} + X_i^T \beta_{j_i}}} \\
Y_i &\sim \text{Bern}(\theta_i) 
\end{align*}

where $j \in \{0, 1\}$ indicates an index for pedestrians and cyclist, $r \in \{0, \dots, 3\}$ indexes the region
of the country (Northast, Midwest, South, West), and $j_i$ indexes the $B_j$ coefficient vector according to whether a
particular crash involved a pedestrian or a cyclist. A full accounting of which regions correspond to which states is 
located in the appendix.

We wish to estimate the effects of the variables separately for both pedestrians and cyclists, but because of our 
relatively sparse feature space, there is some concern of wild coefficient values, so for that reason we use the 
hierarchical model to constrain an overall effect of a feature, and then estimate whether that effect differs
between cyclists and pedestrians. 

Next, for each beta coefficient, we will estimate the effect of the safety features and bodyclass using a Laplace 
distribution as our prior. This is equivalent to a LASSO penalty\cite{tibshirani_regression_1996}, and will help
us to control for the sparsity of our feature space. 


\section{Results}

We see strong relationships across region and vulnerable road user type (cyclist or pedestrian). According to our
model, it is generally safer to be a cyclist. The safest places to do so are ...

Many of the results are exactly as one would expect, and partially align with our hypothesis. For example, the 
coefficient on pickup trucks is strong positive for both cyclists and for pedestrians, indicating that these vehicles
are much more likely to kill a pedestrian or cyclist compared to cars (the baseline group). There is also a general 
trend with the weight of the vehicles. Surprisingly, compared to cars, crossovers are less likely to be 
involved in fatal accidents with pedestrians and cyclists. 

-- Weather


-- Safety Features


Also somewhat surprisingly, the hierarchical specification 
reveals that the impact of body cl


\section{Discussion}

-- Strength of coefficients

-- Implications

\section{Conclusion}

-- Sum up

\section{Support}

Trace-plots

Prior predictive checks

\printbibliography

\end{document}



%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
